---
editor_options: 
  chunk_output_type: console
---

# Exploring microclimate data  

<I would add a description of what this R Script intends to achieve and do>

## Install required libraries
```{r}
library(stringr)
library(dplyr)
library(ggridges)
library(viridis)
library(lubridate)
library(hrbrthemes)
library(ggplot2)
library(lubridate)
library(mgcv)
library(ggplot2)
library(ggridges)
library(viridis)

# read the file
# I would rename the file so that it indicates what data it is
microclimate <- read.csv("data/merged_dt_final.csv")
```

<Write a description here of what that dataset contains?>
<We may also need a script previously to indicate how you merged/cleaned microclimate data>

## Cleaning the combined dataset on microclimate and bird detections  

The above dataset contains both microclimate and time-stamped detections of birds associated with when a microclimate recording was logged. 

```{r}
# convert text to lowercases
names(microclimate) <- tolower(names(microclimate)) 

# verify if the column names are appropriate
names(microclimate)

# fix certain columns
names(microclimate)[1] <- "temp_c"
names(microclimate)[7] <- "true_positive"

# exclude rows with missing temp or time
microclimate <- microclimate[is.finite(microclimate$temp_c) & !is.na(microclimate$time_observation), ]

# filter BirdNET detections > 0.7
# something wrong here - I thought we were filtering detections separately for both species. Can we rewrite the code below to indicate that you are choosing a different confidence threshold for green and the greenish warbler. 

microclimate <- microclimate %>%
  filter(confidence >= 0.7)

## veryify @Alayna or @Eliza if you are left with only ~6000 observations??

#storing date and time ... ex.20240118_083000
x <- c(microclimate$time_observation)
#parse_date_time(x,"Ymd_HMS", tz = "GMT" ), H - mil. time
microclimate$time_observation <- parse_date_time(x,"Ymd_HMS", tz = "GMT" )

#I'll separate graphs and results by double hashtags '##'

##temp and time graphed  
fig_temp_vs_time <- ggplot(microclimate, aes(x = time_observation, y = temp_c, color = factor(elevation))) + 
  geom_line() + theme_minimal() + labs(title = "Temperature through time")

##temp n time w/ facet
# This is a cool plot - one idea is to potentially replot this with regression lines added to see if there is a linear increase over time. It appears that at lower elevations, the temperature is increasing between Feb to April, but at high elevations, there isn't a significant increase??


ggplot(microclimate, aes(x = time_observation, y = temp_c, color = factor(elevation))) + geom_smooth(method = "lm", se = FALSE) +
  geom_line() + facet_wrap( ~ elevation) + theme_minimal() + 
  labs(title = "Temperature through time")
```

## Annotate this section to indicate what the plots are 

```{r}
## This plot is also very interesting - I wonder if it sparks any ideas on how much variation there is in microclimate at low vs. high elevations and if this variation is something that the green warbler can handle while the greenish cannot? Or is it simply resource dependent?

##claus o. wilke - credit: r graph gallery
fig_temp_by_elevation <- 
ggplot(microclimate, aes(x = temp_c, y = factor(elevation), fill = stat(x))) +
  geom_density_ridges_gradient(scale = 2, rel_min_height = 0.01) +
  scale_fill_viridis(name = "temp ", option = "C") +
  labs(title = "temperature distribution by elevation",
       x = "temp",
       y = "elevation band") +
  theme_ipsum() +
  theme(
    panel.spacing = unit(0.1, "lines"),
    strip.text.x = element_text(size = 8), axis.text.y = element_blank()
  )

##violin plot - credit: r graph gallery
fig_temp_variance <- microclimate %>%
  ggplot(aes(x = factor(elevation), y = temp_c, fill = factor(elevation))) +
    geom_violin(trim = FALSE) +
    scale_fill_viridis(discrete = TRUE, alpha = 0.6, option = "A") +
    theme_ipsum() +
    theme(
      legend.position = "none",
      plot.title = element_text(size = 11), axis.text.x = element_blank()
    ) +
    ggtitle("temperature variance by elevation bands") +
    xlab("elevation band") +
    ylab("temp")

##best visualization for me
## Can you reannotate this?? 
ggplot(micro_filter, aes(x = factor(elevation), y = temp_c, fill = factor(elevation))) + geom_boxplot() + theme_minimal() + theme(axis.text.x = element_blank()) #must use factor for elevation (discrete x, continuous y)

```

## Modeling temperature and elevation

```{r}

##relationship btwn elevation and temperature - insightful
model_elev_temp <- lm(temp_c ~ elevation, data = microclimate)
summary(model_elev_temp)

##relationship btwn time and temperature - insightful (prev. said not, oops!)
# I would think about the model formulation here. It may not make sense to convert time to numeric, since the numbers won't mean much - they are technically ~30-minute bins right? So a factor/categorical data might be more relevant. If my interpretation is right, you would expect the model to show an association between temperature and increasing time of day - hotter later in the day.
# I would also ask you both a question of why you are plotting this data?

model_temp_time <- lm(temp_c ~ as.numeric(time_observation), data = microclimate)
summary(model_temp_time)

```


```{r}
#work for the GREEN warbler 
# I think we may need to set different confidence thresholds before we rerun the below section of code. Does that make sense?

#green abundance 
green_micro <- micro_filter %>%
   filter(
  str_detect(common.name, regex("green", ignore_case = TRUE)))

#View(green_micro)

#detecting counts 
abundance_green <- green_micro %>%
  group_by(time_observation, temp_c, elevation) %>%
  summarise(detections = n(), .groups = "drop") %>%
  arrange(time_observation)

#View(abundance_green)

##temp n green w/o facet 
ggplot(abundance_green, aes(x = temp_c, y = detections, color = factor(elevation))) + geom_point() +
   theme_minimal() + labs(title = "Temperature v green detections")

##temp n green w/ facet 
ggplot(abundance_green, aes(x = temp_c, y = detections, color = factor(elevation))) + geom_point() + facet_wrap( ~ elevation) + theme_minimal() + labs(title = "Temperature v green detections")


gam_model <- gam(detections ~ s(temp_c) + s(elevation), data = abundance_green)
summary(gam_model)

```

```{r}
#work for the GREENISH warbler 

#greenish abundance 
greenish_micro <- micro_filter %>%
   filter(
    str_detect(common.name, regex("greenish", ignore_case = TRUE)))

#detecting counts 
abundance_ish <- greenish_micro %>%
  group_by(time_observation, temp_c, elevation) %>%
  summarise(detections = n(), .groups = "drop") %>%
  arrange(time_observation)

#View(abundance_ish)

##temp n greenish w/o facet 
ggplot(abundance_ish, aes(x = temp_c, y = detections, color = factor(elevation))) + geom_point() +
   theme_minimal() + labs(title = "Temperature v greenish detections")

##temp n greenish w/ facet 
ggplot(abundance_ish, aes(x = temp_c, y = detections, color = factor(elevation))) + geom_point() + facet_wrap( ~ elevation) + theme_minimal() + labs(title = "Temperature v greenish detections")

gam_model2 <- gam(detections ~ s(temp_c) + s(elevation), data = abundance_ish)
summary(gam_model2)

```

```{r}

#joint species info

#are there any temperature preferences?
 
ggplot(micro_filter, aes( x = common.name, y = temp_c, fill = common.name)) + geom_boxplot(alpha = 0.6) + scale_fill_viridis(discrete = TRUE, option = "A") + theme_minimal() + labs(title = "temperature by species")

t.test(temp_c ~ common.name, data = micro_filter)
```







